---
title: "CSC3107 Tidyverse Lab: ASEAN Per-Capita Meat Production"
subtitle: "Group Assignment 3"
date: "2026-05-19"
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  message: false
  warning: false
---

## Setup

Only `tidyverse`, `readxl`, and `waldo` are loaded. The workbook and reference files are read with project-relative paths.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(readxl)
library(waldo)
```

## Team Roles

- `Joshua` - Coder: typed and edited the canonical QMD, kept file paths project-relative, and checked that code chunks run.
- `Benjamin` - Navigator: tracked the task requirements, checked that all required questions were attempted, and guided the transformation strategy.
- `MeiYuen` - Verifier: checked row counts, column classes, units, missing values, join behavior, and reference comparisons.
- `Dannon` - Constructive Critic + Reporter: reviewed the report structure, explanations, style, and final submission checklist.
- `Joshua` - AI Liaison: kept the AI interaction log and reviewed AI suggestions against the assignment rules.

No late arrivals or role changes were recorded.

## 3.1 Think Before Prompting

Before using AI, the team expected the tidy meat table to have one row for each country, animal type, and year. The columns and classes should be:

| Column | Class | Example value |
|---|---|---|
| `country` | character | `"Brunei Darussalam"` |
| `animal` | character | `"beef and buffalo"` |
| `year` | numeric | `2021` |
| `production` | numeric | `4.26741` |

The expected number of rows is `10 countries * 4 animal types * 4 years = 160` rows.

Sheet `X.14` is not tidy because years are stored as section rows such as `2021`, `2020`, `2019`, and `2018` instead of values in a `year` column. Countries are spread across many columns, so the country variable is encoded in column headers. The row labels also mix real animal categories with grouping and subtotal rows such as `Livestock production`, `Poultry production`, and `Total meat production`. The sheet also includes presentation material such as title, unit, source, and note rows near the data.

The team predicted that poultry and pig meat would dominate many per-capita values because these are common meats and can usually be produced at larger scale than cattle, sheep, or goats. We also expected Singapore to have low domestic production per resident because of its limited land area and dependence on food imports.

## 3.2 Import the ASEAN Meat Production Data

The relevant meat production table is on sheet `X.14`. The selected range keeps the country-level table and ASEAN aggregate column while excluding title rows and source notes.

```{r}
#| label: import-meat

meat_raw <- read_xlsx(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "X.14",
  range = "A4:L36"
)

meat_raw
```

## 3.3 Tidy the Meat Production Data

The year rows are filled down, subtotal rows are removed, the ASEAN aggregate column is dropped, and the country columns are pivoted into one `country` column.

```{r}
#| label: tidy-meat-before-correction

meat_tidy_uncorrected <- meat_raw |>
  rename(animal = `Animal Type`) |> # Renames Animal Type to animal
  select(-ASEAN) |>                 # Drop ASEAN column

  # Creates new column `year`
  mutate(year = as.numeric(str_extract(animal, "^\\d{4}$"))) |>
  fill(year) |>
  # Drop all rows under animal that does not match the following:
  filter(animal %in% c(
    "Beef and Buffalo Meat",
    "Mutton and Goat Meat",
    "Pig meat",
    "Poultry meat"
  )) |>
  # Convert all remaining columns except animal and year and turns them into country and production 

  pivot_longer(
    cols = -c(animal, year),
    names_to = "country",
    values_to = "production"
  ) |>

  # Convert animal values to lower-case and match new states
  mutate(
    animal = recode(
      animal,
      "Beef and Buffalo Meat" = "beef and buffalo",
      "Mutton and Goat Meat" = "mutton and goat",
      "Pig meat" = "pig",
      "Poultry meat" = "poultry"
    ),
    production = as.numeric(production)
  ) |>
  # Sorting rows in descending order by year, breaking ties with animal and country
  select(country, animal, year, production) |>
  arrange(desc(year), animal, country)

print(head(meat_tidy_uncorrected))
print(tail(meat_tidy_uncorrected))
```

> Verifier Check: the tidy table has 160 rows, four columns, and exactly the four required animal labels.

```{r}
#| label: verify-meat-shape-and-animals

meat_tidy_uncorrected |>
  summarise(
    rows = n(),
    columns = ncol(meat_tidy_uncorrected),
    animals = str_c(sort(unique(animal)), collapse = ", ")
  )
```

## 3.4 Correct Cambodia's Mutton-and-Goat Entries

Cambodia's mutton-and-goat entries are coded as `0` in the XLSX export, but the PDF source marks those values as unavailable. Treating unavailable data as true zero production would bias the per-capita result downward, so the correction is limited to the four affected observations: Cambodia, `mutton and goat`, and years 2018 through 2021.

```{r}
#| label: correct-cambodia-mutton-goat

meat_tidy <- meat_tidy_uncorrected |>
  mutate(
    production = if_else(
      country == "Cambodia" &
        animal == "mutton and goat" &
        year %in% 2018:2021 &
        production == 0,
      NA_real_,
      production
    )
  )

# Replaces all 0 elements under Cambodia's mutton and goat section with NA, else we retain the original production value

meat_tidy |>
  filter(country == "Cambodia", animal == "mutton and goat")
```

> Verifier Check: the verifier used a different approach with the `$` operator and compared the result to the Coder's corrected table.

```{r}
#| label: verify-cambodia-correction

meat_tidy_verifier <- meat_tidy_uncorrected
meat_tidy_verifier$production[
  meat_tidy_verifier$country == "Cambodia" &
    meat_tidy_verifier$animal == "mutton and goat" &
    meat_tidy_verifier$year %in% 2018:2021 &
    meat_tidy_verifier$production == 0
] <- NA_real_

compare(meat_tidy, meat_tidy_verifier, tolerance = 1e-6)
```

The first and last six rows after correction confirm the requested ordering.

```{r}
#| label: show-meat-tidy

print(head(meat_tidy))
print(tail(meat_tidy))
```

## 3.5 Checkpoint: Compare `meat_tidy` to Reference Data

```{r}
#| label: compare-meat

meat_ref <- read_csv("meat_tidy_reference.csv", show_col_types = FALSE)

compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

## 3.6 Tidy the Population Data

The population table is on sheet `I.1`. The range excludes the ASEAN aggregate row.

```{r}
#| label: import-population

pop_raw <- read_xlsx(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  range = "A5:K15"
)

pop_raw
```

```{r}
#| label: tidy-population

pop_tidy <- pop_raw |>
  rename(country = Country) |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "pop"
  ) |>
  mutate(
    year = as.numeric(year),
    pop = as.numeric(pop)
  ) |>
  select(country, year, pop) |>
  arrange(desc(year), country)

pop_tidy
```

> Verifier Check: if `trim_ws = FALSE`, the raw workbook preserves the trailing space in `"Malaysia "`. The default `read_xlsx()` setting is `trim_ws = TRUE`, so `pop_tidy$country` contains `"Malaysia"` without the trailing space.

```{r}
#| label: verify-malaysia-trailing-space

pop_raw_no_trim <- read_xlsx(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  range = "A5:K15",
  trim_ws = FALSE
)

pop_raw_no_trim |>
  filter(str_detect(Country, "Malaysia")) |>
  transmute(country = Country, characters = nchar(Country))

pop_tidy |>
  filter(country == "Malaysia") |>
  distinct(country) |>
  mutate(characters = nchar(country))
```

## 3.7 Checkpoint: Compare `pop_tidy` to Reference Data

```{r}
#| label: compare-population

pop_ref <- read_csv("population_tidy_reference.csv", show_col_types = FALSE)

compare(pop_tidy, pop_ref, tolerance = 1e-6)
```

## 3.8 Join Meat Production to Population

A left join preserves every row and the row order of `meat_tidy` while adding population values for matching countries and years.

```{r}
#| label: join-meat-population

merged <- meat_tidy |>
  # Using left_join to preserve every row of meat_tidy
  left_join(pop_tidy, by = c("country", "year")) |>
  select(country, animal, year, production, pop)

merged
```

> Verifier Check: the ordering of the joined table is the same as the ordering of `meat_tidy`.

```{r}
#| label: verify-merged-row-order

identical(
  merged |> select(country, animal, year, production),
  meat_tidy
)
```

## 3.9 Derive Per-Capita Production

Both source variables are stored in thousands, so kilograms per person is calculated as `production / pop * 1000`.

```{r}
#| label: derive-meat-per-capita

merged <- merged |>
  mutate(meat_per_cap = production / pop * 1000)

merged
```

> Verifier Check: the verifier independently recalculated `meat_per_cap` with the `$` operator.

```{r}
#| label: verify-meat-per-capita

merged_verifier <- merged |>
  select(-meat_per_cap)

merged_verifier$meat_per_cap <-
  merged_verifier$production / merged_verifier$pop * 1000

compare(merged$meat_per_cap, merged_verifier$meat_per_cap, tolerance = 1e-6)
```

## 3.10 Checkpoint: Compare `merged` to Reference Data

```{r}
#| label: compare-merged

merged_ref <- read_csv("merged_reference.csv", show_col_types = FALSE)

compare(merged, merged_ref, tolerance = 1e-6)
```

## 3.11 Visualize the Result

```{r}
#| label: visualize-meat-per-capita

ggplot(
  merged,
  aes(x = year, y = meat_per_cap, color = animal)
) +
  geom_line() +
  geom_point() +
  labs(
    x = NULL,
    y = "Production (kg per person)",
    color = "Animal",
    title = "Per-Capita Meat Production in ASEAN, 2018-2021",
    caption = "Source: ASEAN Statistical Yearbook 2023"
  ) +
  xlim(2017.9, 2021.1) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1", "10", "100")
  ) +
  facet_wrap(~country) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification.inside = c(1, 0.05)
  )
```

## 3.12 Comment on the Plot

The plot should be read as domestic production per resident, not direct meat consumption, because imports and exports are not included. A country can therefore show low production per person even if residents consume imported meat. Also, because the y-axis is log-scaled, the discussion below uses the calculated `meat_per_cap` values rather than judging the visual slope alone. The values are rounded to one decimal kilogram per person. A change can come from a change in domestic production, a change in population, or both.

Across the available animal categories, poultry and pig production explain most of the visible increases and decreases. This is consistent with broader meat-market evidence from the OECD-FAO Agricultural Outlook 2025-2034, which says productivity gains have been greater in pig and poultry production than in cattle and sheep, and that lower relative prices support demand for non-ruminant meats.^[OECD and FAO. "Meat." *OECD-FAO Agricultural Outlook 2025-2034*. <https://www.oecd.org/en/publications/oecd-fao-agricultural-outlook-2025-2034_601276cd-en/full-report/meat_5462e384.html>] African swine fever (ASF) is also important background for pig production because FAO describes ASF as a major threat in East and Southeast Asia, where many pigs are kept in low-biosecurity smallholder systems.^[FAO ECTAD Asia and the Pacific. "African swine fever." <https://www.fao.org/ectad/asiapacific/our-focus/african-swine-fever/en>] In this dataset, the largest overall increase from 2018 to 2021 is Brunei Darussalam, while the largest overall decrease is Myanmar. Cambodia's total comparison excludes the corrected mutton-and-goat observations because those values are unavailable.

- `Brunei Darussalam`: Brunei has the clearest increase. Total available meat production per resident rises from about 59.1 kg/person in 2018 to 83.6 kg/person in 2021, an increase of 24.6 kg/person. Poultry remains the dominant category and rises from 57.2 to 73.5 kg/person, while beef-and-buffalo also increases sharply from 1.6 to 9.9 kg/person. External sources support a production-capacity explanation: Brunei's Department of Agriculture and Agrifood reported higher broiler output in the first half of 2021 because more farm land and poultry houses were added, and the same report linked higher local beef/buffalo slaughter to constrained frozen imports and the use of live imports for local slaughter.^[Biz Brunei. "Local chicken supply up 25% in first half of 2021." <https://www.bizbrunei.com/2021/08/local-chicken-supply-up-25-in-first-half-of-2021/>] Brunei's agriculture statistics also report high broiler self-sufficiency, so the poultry-led increase is consistent with a domestic food-security and self-sufficiency push rather than a random fluctuation.^[Department of Agriculture and Agrifood, Brunei Darussalam. *Brunei Darussalam Agriculture Statistics in Brief 2021*. <https://www.agriculture.gov.bn/SiteCollectionDocuments/Statistik/Brunei%20Darussalam%20Agriculture%20Statistics%20in%20Brief%202021.pdf>]
- `Cambodia`: Cambodia is comparatively stable. Total available production moves only from about 11.3 to 11.7 kg/person between 2018 and 2021. Pig increases from 5.4 to 6.5 kg/person, but beef-and-buffalo decreases from 4.2 to 3.6 kg/person. Poultry stays almost flat, moving from 1.7 to 1.7 kg/person after a small rise in 2019 and a small dip in 2020. The mutton-and-goat line is missing because those four observations were corrected to `NA`, so Cambodia should not be interpreted as having zero mutton-and-goat production. A likely reason for the mixed movement is post-ASF adjustment: FAO says ASF was first detected in Cambodia in April 2019 and that pig production declined compared with 2018, while later Cambodian reporting pointed to growth in animal and livestock production value in 2021.^[FAO Cambodia. "FAO strengthens field capacities of Cambodia's officials to fight African swine fever (ASF)." <https://www.fao.org/cambodia/news/detail/FAO-strengthens-field-capacities-of-Cambodia-s-officials-to-fight-African-swine-fever-%28ASF%29-/en>]^[Agence Kampuchea Presse. "Animal Production Valued at US$5.1 Billion in 2021." <https://www.akp.gov.kh/post/detail/245734>] This supports a cautious interpretation: the pig increase in the table may reflect recovery from earlier ASF disruption, while poultry and beef did not move enough to change the total very much.
- `Indonesia`: Indonesia shows a moderate net increase, from about 13.6 to 17.6 kg/person. This increase is mainly caused by poultry, which rises from 9.8 to 14.3 kg/person. The poultry increase is strongest between 2018 and 2019, then it decreases in 2020 and partly recovers in 2021. Beef-and-buffalo declines from 2.1 to 1.7 kg/person, while pig falls in 2019 and recovers by 2021. The 2020 poultry dip is consistent with COVID-era demand and supply pressure: USDA reporting on Indonesia's grain and feed sector links food-service closures, weaker purchasing power, and feed dependence to pressure on livestock and poultry demand.^[USDA Foreign Agricultural Service. *Grain and Feed Annual: Indonesia, 2021*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Grain+and+Feed+Annual_Jakarta_Indonesia_03-27-2021.pdf>] The 2021 partial recovery is therefore plausibly connected to demand normalization after the initial COVID shock, but the report does not prove that poultry alone caused the total increase.
- `Lao PDR`: Lao PDR has a steady upward pattern across all four animal categories. Total available production rises from about 22.3 to 28.5 kg/person. Pig is the largest contributor, increasing from 10.0 to 13.8 kg/person. Poultry also increases from 4.4 to 5.9 kg/person, and beef-and-buffalo rises from 7.6 to 8.4 kg/person. Since each year from 2018 to 2021 is higher than the previous year in the total, Lao PDR has one of the most consistent incremental trends. The likely reason is sector expansion rather than a one-year shock: Lao PDR reporting to the UNFCCC uses national agriculture statistics showing livestock inventories rising around 2020 to 2021, and a later ILO livestock-sector review describes livestock as a growth area supported by commercialisation and government targets.^[Lao PDR. *Biennial Transparency Report / National Inventory Report*. <https://unfccc.int/sites/default/files/resource/3.%20NIR_BTR1_Lao%20PDR.pdf>]^[International Labour Organization. *Commercializing livestock production in Lao PDR*. <https://www.ilo.org/sites/default/files/2024-09/PDR_Livestock_C_18_FINAL_WEB.pdf>] ASF remained a regional risk for pigs, including Lao PDR, so the steady pig rise should still be read with disease-risk context rather than as guaranteed long-term stability.^[FAO Newsroom. "One year on, close to 5 million pigs lost to Asia's swine fever outbreak." <https://www.fao.org/newsroom/detail/One-year-on-close-to-5-million-pigs-lost-to-Asia-s-swine-fever-outbreak/ar>]
- `Malaysia`: Malaysia remains high in poultry production per resident, but it decreases overall from about 65.3 to 58.0 kg/person. The main decrement is poultry, which falls from 57.8 to 49.9 kg/person. This decline is partly offset by pig, which increases from 5.7 to 6.7 kg/person. The total falls in 2019 and 2020, then has a small recovery in 2021. Malaysia therefore still has a high poultry level, but the direction over the period is downward rather than upward. External sources suggest several pressures behind this movement. DOSM reported that several livestock products, including beef, mutton, and pork, declined in 2021, while other reporting based on DOSM linked 2020 food-output declines to farm downsizing and labour shortages during COVID-19.^[Department of Statistics Malaysia. "Selected Agricultural Indicators, Malaysia, 2021." <https://www.dosm.gov.my/portal-main/release-content/selected-agricultural-indicators-malaysia-2021>]^[Malay Mail. "DoSM: Covid-19 pandemic cut Malaysia's 2020 food output as farms downsized, fish catch fell from foreign labour shortage." <https://www.malaymail.com/news/malaysia/2021/08/26/dosm-covid-19pandemic-cut-malaysias-2020-food-output-as-farms-downsizedfi/2000670>] For poultry, higher feed costs are a plausible additional pressure because Malaysia relies heavily on imported corn and soybean meal for poultry feed, and the Ministry of Finance noted that higher imported feed prices affected local chicken breeders.^[Ministry of Finance Malaysia. "MOF to study best method to stabilise chicken prices." <https://www.mof.gov.my/portal/en/news/press-citations/mof-to-study-best-method-to-stabilise-chicken-prices>]
- `Myanmar`: Myanmar has the largest decrease in the plot. Total available production falls from about 65.5 to 20.2 kg/person, mostly because the 2021 values are much lower than the 2020 values. Poultry drops from 35.2 to 12.3 kg/person, pig drops from 19.7 to 5.1 kg/person, and beef-and-buffalo drops from 8.7 to 2.5 kg/person. Since most of the decrease happens between 2020 and 2021, Myanmar should be discussed as a sharp late-period fall in reported domestic production per resident, not as a gradual four-year decline. The timing is consistent with severe 2021 disruption. The World Bank described Myanmar's economy as being hit by the military coup and COVID-19, with logistics, electricity, connectivity, exchange-rate, and input-cost constraints.^[World Bank. "Myanmar Economic Monitor January 2022: Contending with Constraints." <https://www.worldbank.org/en/country/myanmar/publication/myanmar-economic-monitor-january-2022-economic-activity-in-myanmar-to-remain-at-low-levels-with-the-overall-outlook-blea>] USDA reporting is more directly related to livestock: it reported reduced live-bird supply into Yangon, transport limits, feed-mill labour shortages, and sharply lower feed production after the coup.^[USDA Foreign Agricultural Service. *Burma - Military Coup Continues to Hamper Agricultural Trade*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Burma+-+Military+Coup+Continues+to+Hamper+Agricultural+Trade_Rangoon_Burma+-+Union+of_04-12-2021.pdf>] These sources make the 2021 decline credible as a production and supply-chain shock, especially for poultry and pigs.
- `Philippines`: The Philippines shows a steady overall decrement, from about 34.3 to 25.1 kg/person. Pig is the main reason for the fall, decreasing from 17.7 to 10.8 kg/person. Beef-and-buffalo also decreases from 2.9 to 1.7 kg/person, while poultry is comparatively stable but slightly lower by 2021. The total decreases each year, so the Philippines is one of the clearer examples of a consistent downward trend across the period. The strongest explanation is ASF in the swine sector. The Philippine Statistics Authority reported that hog output in the first quarter of 2021 fell 25.8% year-on-year and swine inventory fell 22.6%, while the Department of Agriculture linked the pork-supply slump to ASF, movement restrictions, feed, and shipping issues.^[Philippine Statistics Authority. *Swine Situation Report, January-March 2021*. <https://psa.gov.ph/system/files/livestock-poultry/6_%255BONSrev-cleared%255D%2520170521%2520SR_Q1%25202021%2520Swine%2520Situation%2520Report_signed.pdf>]^[Department of Agriculture, Philippines. "OneDA Family 2021 Yearender: Propping up corn production, controlling ASF to revive livestock, poultry sector." <https://www.da.gov.ph/oneda-family-2021-yearender-propping-up-corn-production-controlling-asf-to-revive-livestock-poultry-sector/>] Poultry did not offset the pig decline because COVID restrictions also weakened demand from restaurants, processors, and institutional buyers, according to USDA and Philippine government summaries.^[USDA Foreign Agricultural Service. *2021 Livestock and Poultry Annual: Philippines*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=2021+Livestock+and+Poultry+Annual_Manila_Philippines_08-30-2021.pdf>]
- `Singapore`: Singapore remains low for beef-and-buffalo and mutton-and-goat, with both close to zero throughout the period. Its total available production changes only slightly, from about 22.2 to 21.3 kg/person, but the composition changes. Poultry decreases from 18.9 to 15.6 kg/person, while pig increases from 3.3 to 5.7 kg/person. This means the decrease in poultry is partly offset by the increase in pig. The reason is structural: Singapore has limited land and imports more than 90% of its food, so domestic production is not expected to track domestic consumption closely.^[Ministry of Sustainability and the Environment. "Food." <https://www.mse.gov.sg/policies/food>] Singapore Food Agency statistics similarly state that only about 1% of land is available for food production and that local production policy focuses on resilience rather than full self-sufficiency in conventional meat.^[Singapore Food Agency. *Singapore Food Statistics 2021*. <https://www.sfa.gov.sg/docs/default-source/publication/sg-food-statistics/singapore-food-statistics-2021.pdf>] The small pig increase and poultry decrease should therefore be read as changes in a narrow domestic production base, not as evidence that Singaporeans shifted their meat consumption in the same direction.
- `Thailand`: Thailand is one of the most stable countries in the plot. Total available production increases slightly from about 43.2 to 44.9 kg/person. Poultry increases from 26.2 to 28.3 kg/person and remains the largest category. Pig decreases slightly from 14.7 to 14.2 kg/person, while beef-and-buffalo fluctuates but ends close to its starting value. The pattern is therefore a small positive increment, mainly from poultry, rather than a major structural change. USDA reporting on Thailand's poultry sector expected chicken production to increase slightly in 2021, supported by domestic consumption and exports, but also noted constraints from COVID outbreaks at processing plants, weak tourism and food-service demand, and high feed costs.^[USDA Foreign Agricultural Service. *Poultry and Products Annual: Thailand, 2021*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Poultry+and+Products+Annual_Bangkok_Thailand_09-01-2021.pdf>] This explains why Thailand's poultry line rises only gradually rather than sharply. The pig line should be interpreted cautiously: Thailand officially confirmed ASF after the period, in January 2022, but USDA later reported that pork supplies had already been disrupted and that production had been reduced by disease-control and feed-cost pressures.^[USDA Foreign Agricultural Service. *Thai Pork Supplies Disrupted by ASF*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Thai+Pork+Supplies+Disrupted+by+ASF_Bangkok_Thailand_01-14-2022.pdf>]
- `Viet Nam`: Viet Nam decreases overall from about 54.5 to 47.3 kg/person, but the animal categories move in opposite directions. Pig decreases strongly from 40.0 to 26.3 kg/person, while poultry increases from 9.9 to 17.3 kg/person. This means the poultry increment offsets part of the pig decrement, but not enough to prevent an overall decline. Viet Nam increases from 2019 to 2020, then falls in 2021 because the pig decrease is larger than the poultry increase. The main reason is the legacy of ASF and uneven recovery in swine production. Viet Nam's National Statistics Office states that pig farming was heavily affected by ASF in 2019, with pig live-weight output falling compared with 2018, followed by partial recovery in 2020 and 2021.^[National Statistics Office of Viet Nam. "Livestock of pig has recovered." <https://www.nso.gov.vn/en/data-and-statistics/2021/04/livestock-of-pig-has-recovered/>] USDA reporting also notes that millions of pigs were culled or died after ASF, and that recovery was led by larger industrial operations.^[USDA Foreign Agricultural Service. *Vietnam African Swine Fever Update, 2021*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Vietnam+African+Swine+Fever+Update_Hanoi_Vietnam_05-08-2021.pdf>] Poultry rose partly because poultry production was encouraged as an alternative protein source while the swine sector recovered, but later 2021 COVID restrictions, slaughterhouse disruptions, and high stocks also created poultry-sector pressure.^[USDA Foreign Agricultural Service. *Oilseeds and Products Annual: Vietnam, 2021*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Oilseeds+and+Products+Annual_Hanoi_Vietnam_04-01-2021.pdf>]^[USDA Foreign Agricultural Service. *Oilseeds and Products Update: Vietnam, 2021*. <https://apps.fas.usda.gov/newgainapi/api/Report/DownloadReportByFileName?fileName=Oilseeds+and+Products+Update_Hanoi_Vietnam_10-01-2021>]

Overall, the plot suggests that country-level changes are not uniform across meat types or across causes. Brunei Darussalam and Lao PDR show increases that are consistent with livestock-capacity expansion and food-supply policy. Indonesia and Thailand show smaller poultry-led increases, but both are moderated by COVID-era demand, processing, and feed-cost pressures. Myanmar and the Philippines show clearer disruption stories: Myanmar's sharp 2021 decline is consistent with political, logistics, and feed-supply shocks, while the Philippines' decline is strongly linked to ASF in pigs. Malaysia and Viet Nam show mixed structures where poultry and pig move in different directions, but the overall result is downward. Singapore is a special case because its domestic production is structurally limited by land scarcity and import dependence. Cambodia is almost flat once the unavailable mutton-and-goat values are treated as missing.

## 4.1 Reflect on AI Use

Tool used: ChatGPT Desktop App.

One useful AI suggestion was to use `read_xlsx()` ranges that start at the actual table headers, then use `pivot_longer()` after filling the year rows. This was accepted after checking the workbook and confirming that the resulting `meat_tidy`, `pop_tidy`, and `merged` tables matched the three reference CSV files.

One AI limitation was that an early draft suggested wrapping `compare()` output in a custom `"No differences found."` message. That was revised because the task sheet specifically asks for direct `waldo::compare()` checkpoints that print the standard no-differences message.

This reflection was checked against the available `ai_log.md` before submission. The team should replace the reconstructed response text in the log if a full exported transcript becomes available.

## 4.5 Packaging Checklist

- The final QMD renders without errors.
- The rendered HTML file is included.
- `ai_log.md` is included and has been reviewed against the actual chat transcript.
- `ASEAN-Statistical-Yearbook-2023.xlsx` is included.
- `meat_tidy_reference.csv`, `population_tidy_reference.csv`, and `merged_reference.csv` are included.
- All file paths in the QMD are project-relative.
- Every code chunk has a unique label.
- The three `compare()` checkpoints show no differences.
- The final archive is named `team-SteelBlue-tidyverse-lab.zip`, using the team identifier recorded in the AI log.
